Data Migration under a Changing Schema in Ampersand

### 2023-10-07 2022Migration/articleMigrationFACS.tex
minor, to avoid an overfull box
violations

This paper defines a migration process that preserves the semantics of data by means of rules (constraints).
The migration process is based on assumptions and requirements that aim to capture a large class of data migrations in practice.
This paper focuses on correctness of the migration. Efficiency of the migration is outside its scope.

\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be upgraded at an ever-increasing pace to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly upgrades of existing systems.
   If an upgrade does not change the schema, we don't need a data migration.
   Yet, schema changes cannot always be avoided, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   So, development teams will try to avoid schema changes at all costs, fearing the risk and effort of the data migration,
   and for a good reason.

   Data migration for other purposes than schema change has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers can and will avoid schema changes and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
   Another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014}, defines a migration as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   This does not cater for schema changes, however.
   Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
   If the schema is not available at compile time, the work will have to be done at runtime.
   This leads to versioned storage and an overhead in performance.
   This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
   The prototypical use case for that is to frequently release upgrades of information systems in production,
   where the semantic integrity of data must be preserved across schema changes.
   Another use case is application integration for multiple dispersed data sources with explicit schemas.

   To analyze data migrations with changing schemas, we need to define the notion of schema and how it constrains the data.
   To facilitate a formal analysis of the situation,
   we define an information system (in section~\ref{sct:Information Systems}) as a pair of a schema and a dataset,
   so that we have the schema explicitly at our disposal.
   The dataset is separate, so we may consider the analysis of just a schema as a compile-time activity.
   Section~\ref{sct:Schemas} formalizes schemas as a combination of concepts, relations and rules.
   Relations are carriers of the data and concepts are needed to define them.
   Rules constrain the data to impose semantics on the data.
   These constraints must be satisfied all the time, so we call them {\em invariant}.
   For experimental validation we have used the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
   because its syntax and semantics correspond directly to these definitions.
   Ampersand has a notion of rule, which implements the invariants on the dataset.
   In this way, prototypes built in Ampersand have helped to refine the theory.

   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, invariants that change yield extra work for software engineers to write the transform code,
   for which ETL tools typically provide little support.
   The pressure on increasing deploy times calls for further automation of SCDMs.
   Hence, we can expect more frequent SCDMs, and downtime for the sake of migration becomes less acceptable.
   So, our research aims for zero-downtime.
   Another practical problem is that of data quality.
   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean up data.
   Some of that must be done before the migration; some can wait till after the migration.
   We can define part of the data quality as satisfying semantic constraints.
   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
   a registration of valid addresses can be checked automatically.
   An information system that is defined by constraints on a data set
   can signal this type of data pollution by adding such constraints.
   Some forms of data pollution cannot be detected in this way, however.
   An example is when a person has deliberately specified a false name without violating any conceivable constraint.

   To make the case for zero-downtime, we must distinguish between two types of constraint:
\begin{enumerate}
\item Invariant\\
   An \define{invariant} is a constraint that is always true in a system.
   This corresponds to the classical database transaction.
   Violation of an invariant means it has to be restored without the outside world noticing.
   This requires either human intervention or an automated procedure.
   If it is restored, the system commits to the new state, which satisfies the constraint.
   If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
   In both cases, the invariant remains true.
\item Business constraint\\
   A \define{business constraint} is a constraint that can be violated temporarily until a user restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   We consider business constraints not to be invariants because users can violate them.
   They are not true all the time.
\end{enumerate}
   Suppose some invariant, $u$, of the {\em desired system} does not apply to the {\em existing system},
   and suppose it would be violated if applied to the data of the existing system.
   Now, when we spin up the desired system on the data set of the existing system, we have a problem.
   Every invariant of the desired system must be satisfied but $u$ isn't.
   So we cannot spin it up.
   In many cases, clearing out all violations before spinning up the desired system might take too much time and new violations will arise in the process.
   To solve this problem, we implement every invariant in the migration system initially as a business constraint,
   letting users restore invariance while the new system is running.
   For this purpose, we must define an intermediate system, the \define{migration system},
   which implements the invariants of the desired system as business constraints.
   This migration system ensures that every violation that is fixed will not return as a violation.
   In this way, the business constraint gradually turns into an invariant.
   Since the number of violations is finite, it is up to the business to resolve these violations.
   New violations that occur are under control of the desired system,
   so the business experiences the migration system as the desired system.
   In this way, the migration system bridges the gap and users get a zero-downtime SCDM.

   Summarizing, the following requirements apply to SCDMs
\begin{enumerate}
\item users must experience zero-downtime, to enable more frequent SCDMs.
\item users must be able to fix business constraints, so we need an intermediate ``migration system'' to deliver zero-downtime.
\item the number of violations that users must fix is finite and decreases (monotonically) over time,
   to ensure that the migration system can be removed after the migration is done.
\end{enumerate}

@@

   Actors (both users and computers) are changing the data in a system continually.
   The state of the system is represented by a data set, typically represented in a persistent store or database.
   Events that the system detects may cause the state to change.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
   \end{center}
\caption{Anatomy of an information system}
\label{fig:pre-migration}
\end{figure}
   To keep our theory technology independent, data sets are assumed to contain triples.
   This makes our theory valid for any kind of database that triples can represent,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
   An important aspect of the system's semantics are the constraints, sometimes known as invariants or integrity rules.
   %The purpose of these rules is to keep all constraints satisfied.

   Like most software generators, Ampersand is a compiler that generates information systems from a script.
   Ampersand uses a form of heterogeneous relation algebra,
   which works with relations in a way similar to Alloy~\cite{Alloy2006}.
   A system is then described by a set of rules (aka constraint / invariant), which the system keeps satisfied.
   So all constraints are explicitly available as rules in the Ampersand code.

@@

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   %Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
   %
   In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
   while transferring data in a controlled fashion, as shown in Figure~\ref{fig:migration phase}.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{figures/datamigration-Migration-phase.png}
   \end{center}
\caption{Migration phase}
\label{fig:migration phase}
\end{figure}
   To automate the migration as much as possible,
   we also deploy a migration system,
   which has its own schema to specify the migration itself.
   The migration system comprises the schemas of both the existing system and the desired system,
   to implement the copying of data.
   To fully automate every data migration is unrealistic.
   A migration engineer can minimize manual work by minimizing the size of the increments,
   but she cannot eliminate it completely.
   Data pollution, new business rules, or known issues in the existing system
   may occasionally require the migration engineer to tailor the migration script to specific needs.

@@

\section{Terminology}
\label{sct:Terminology}
   An {\em information system} is a combination of data set, schema, and functionality.
   For the purpose of this paper, functionality captured in user interfaces are ignored because it does not impact the migration.
   Section~\ref{sct:Data sets} desribes data sets. Schemas are treated in section~\ref{sct:Schemas}.
   Then section~\ref{sct:Information Systems} defines information systems.

@@

   Relations serve to organize and store data, to allow a developer to represent facts.
   In this paper, variables $r$, $s$, and $d$ represent relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
   All relations are taken from an infinite set $\Rels$.
   $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.
   Every relation $r$ has a name, a source concept, and a target concept.
   The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.
   The part $\pair{A}{B}$ is called the {\em signature} of the relation.

@@

   A relation $r$ can serve as a container of pairs,
   as defined by the function $\popF{r}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
   It defines a set of pairs, also known as the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
\label{eqn:pop-rel}
\end{equation}
%   Equation~\ref{eqn:wellTypedEdge} implies that for every data set $\dataset$:
%\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\instance_{\dataset}A\ \wedge\ b\instance_{\dataset}B\]
%   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
%
   We overload the notation $\popF{}$ so we can use it on concepts $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
   and expressions.
\begin{align}
   \pop{C}{\dataset}&= \{ x\mid\ x\ \instance_{\dataset}\ C\}
\label{eqn:pop-concept}\\
   \pop{x-y}{\dataset}&= \pop{x}{\dataset} - \pop{y}{\dataset}
\label{eqn:pop-expr}
\end{align}

\subsection{Schemas}
\label{sct:Schemas}
   Schemas serve to capture the semantics of a system~\cite{Spivak2012}.
   We consider a schema to represent the static part of a system, defining concepts, relations, and rules.
   A software engineer defines a schema on design time, so that semantic checks can be implemented at compile time.

   We describe a schema $\schema$ via a tuple $\quadruple{\concepts}{\rels}{\rules}{\enforces}$,
   in which $\concepts\subseteq \Concepts$ is a finite set of concepts,
   $\rels\subseteq \Rels$ is a finite set of relations,
   $\rules\subseteq \Triple{(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}})}{\Concepts}{\Concepts}$ is a finite set of rules,
   and $\enforces\subseteq \Pair{\Rels}{\left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)}$ is a finite set of enforcement rules.

   Each rule in a schema serves to constrain the data set at runtime, to ensure its semantic integrity.
   Every rule is an element of an infinite set called $\Rules$.
   In this paper, variables $u$ and $v$ represent rules.
   A rule $u$ in the schema is a triple $\triple{\violC{u}}{A}{B}$ such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
   In Ampersand, rules are well-typed, just like relations:
   Every violation $\pair{a}{b}$ of a rule $\triple{\violC{u}}{A}{B}$ must satisfy $a\instance A$ and $b\instance B$.
   % For every data set $\dataset$, the set $\viol{u}{\dataset}$ is a subset of $\pop{A}{\dataset}\times\pop{B}{\dataset}$.
   If there are no violations, that is: $\viol{u}{\dataset} = \emptyset{}$, then we say that $u$ is satisfied.

@@

   A rule, {\tt EnrollRule} completes the script.
   It states that a student can enroll for any module that is part of a course she takes.
   % In Ampersand, which is a syntactically sugared form of relation algebra~\cite{JoostenRAMiCS2017},
   % each rule has a name and each rule has a role to maintain its invariance:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
\end{verbatim}
   The Ampersand compiler defines $\viol{\tt EnrollRule}{\dataset}$ as
\begin{equation}
   \pop{\tt isEnrolledFor}{\dataset}-\{\pair{s}{m} \mid\exists c.~s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
\end{equation}
   This rule is satisfied if $\viol{\tt EnrollRule}{\dataset}=\emptyset$, which means that
\begin{equation}
   \begin{array}{l}
   \forall \pair{s}{m}\in\pop{\tt isEnrolledFor}{\dataset}.~ \exists c\in\text{\tt Course}~.\\
   s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
   \end{array}
\label{eqn:example isEnrolledFor}
\end{equation}
   In this example, relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty, so rule {\tt EnrollRule} is satisfied.

@@

   The MoC is reached for dataset $\dataset'$ whenever $\pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'}=\emptyset$ for all rules.
   This should happen after a finite number of inserting events: let $I = \{{\tt fixed}_u \mid u\in\rules_{\schema'}\}$ be the relations that keep track of which violations have been fixed, then there is no infinite chain of systems $\migrsys' = \migrsys_0, \migrsys_1,\ldots$ such that $\migrsys_i\xrightarrowdbl[I]{\overrightarrow\rels_{\schema'}}\migrsys_{i+1}$.
   This shows that when users get rid of violations, there is real progress: they can only do this a finite number of times.
   Next we ask ourselves: does this mean we always reach the MoC?

@@

   For this last question, we need to add a condition:
   There needs to be a system that has the desired schema $\schema'$.
   If rules in $\schema'$ are such that they cannot all be satisfied, regardless of the dataset, no such system exists and we have no hope of reaching it.
   We assume that the rules of $\schema'$ are intended to prevent data pollution, such that a dataset $\dataset'$ describing the `real world' would be one that satisfies them.
   So let $\pair{\dataset'}{\schema'}$ be a system, then we can get to it through our migration system.
   To show this, we construct a migration system $\migrsys^C$ satisfying $\pop{{\tt fixed}_u}{\dataset_{\migrsys^C}}=\pop{{\tt viols}_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
   We can do so by simply stating its triples, since the schema is determined by $\migrsys'$:

@@

   This work does not consider what to do about (user) interfaces.
   Instead, it models events by assuming that any change to the dataset can be achieved.
   In practice, such changes need to be achieved through interfaces.
   Most Ampersand systems indeed allow the users of the system to edit the dataset quite freely through the interfaces.
   However, some interfaces may require certain constraints to be satisfied, which means that interfaces of the desired system might break.
   In the spirit of the approach outlined here, we hope to generate migration interfaces that can replace any broken interfaces until the Moment of Transition.
   How to do this is future work.

>>>>

   This paper defines a migration process that preserves the semantics of data by means of invariants (constraints)
   with zero downtime for users.
   The migration process is based on assumptions and requirements that aim to capture a large class of data migrations in practice.
   This paper focuses on correctness of the migration. Efficiency of the migration is outside its scope.
\keywords{Generative software \and Incremental software deployment \and Data migration \and Relation algebra \and Ampersand \and Schema change. \and Invariants \and Zero downtime.}
\end{abstract}

\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be upgraded regularly to keep up with changing requirements in a dynamically evolving environment.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly upgrades of existing systems.
   In upgrades, schema changes cannot always be avoided, so a {\em schema changing data migration} (SCDM) will be neccessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   The risk and effort of such data migrations explains why development teams try to avoid schema changes at all costs.

   Data migration for other purposes than schema change has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers can and will avoid schema changes and functionality changes to avoid introducing new errors in an otherwise error-prone migration process.
   Another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014}, defines a migration as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   This does not cater for schema changes, however.
   Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
   If the schema is not available at compile time, the work will have to be done at runtime.
   This requires versioned storage of production data and an overhead in performance.
   This paper focuses on SCDMs in situations with an explicit schema that changes with the data migration.
   The prototypical use case for that is to frequently release upgrades of information systems in production,
   where the semantic integrity of data must be preserved across schema changes.
   Another use case is application integration for multiple dispersed data sources with explicit schemas.

   To analyze data migrations with changing schemas, we need to define the notion of schema and how it constrains the data.
   To facilitate a formal analysis of the situation,
   we define an information system (in section~\ref{sct:Information Systems}) as a pair of a schema and a data set,
   so that we have the schema explicitly at our disposal.
   The data set is separate, so we may consider the analysis of just a schema as a compile-time activity.
   Section~\ref{sct:Schemas} formalizes schemas as a combination of concepts, relations and invariants.
   Relations are carriers of the data and concepts are needed to define them.
   Invariants are constraints on the data to impose semantics.
   These constraints must be satisfied all the time.
   For experimental validation we have used the language Ampersand~\cite{JoostenRAMiCS2017,Joosten-JLAMP2018}
   because its syntax and semantics correspond directly to these definitions.
   Ampersand has a notion of rule, which implements the invariants on the data set.
   In this way, prototypes built in Ampersand have helped to refine the theory.

   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, invariants that change yield extra work for software engineers to write the transform code,
   for which ETL tools typically provide little support.
   The pressure to decrease deploy times calls for further automation of SCDMs.
   Hence, we can expect more frequent SCDMs, so downtime for the sake of migration becomes less acceptable.
   That is why our research aims for zero downtime.

   Another practical problem is that of data quality.
   Migrations typically suffer from a backlog of deteriorated data, incurring work to clean up data.
   Some of that must be done before the migration; some can wait till after the migration.
   We can capture part of the data quality problem as a requirement to satisfy semantic constraints.
   E.g. the constraint that the combination of street name, house number, postal code, and city occurs in
   a registration of valid addresses can be checked automatically.
   In a formalism like Ampersand, which allows us to express such constraints, we can add constraints for data quality to the schema.
   This allows us to signal the data pollution at runtime.
   Some forms of data pollution cannot be detected in this way, however.
   An example is when a person has deliberately specified a false name without violating any conceivable constraint.

   To make the case for zero downtime, we must distinguish between two types of constraint:
\begin{enumerate}
\item Invariant\\
   An \define{invariant} is a constraint that is always true in a system.
   This corresponds to the classical database transaction.
   Violation of an invariant means it has to be restored without the outside world noticing.
   This requires either human intervention or an automated procedure.
   Once it is restored, the system commits to the new state, which satisfies the constraint.
   If it cannot be restored, the system refuses the transaction and keeps the state as it was before starting the transaction.
   In both cases, the invariant remains true.

   A special case of an invariant is the \define{enforcement rule}.
   As soon as an enforce rule is violated, the system restores it automatically, without human intervention.
\item Business constraint\\
   A \define{business constraint} is a constraint that can be violated temporarily until a user restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   We consider business constraints not to be invariants because users can violate them.
   They are not true all the time.
\end{enumerate}
   Suppose some invariant, $u$, of the {\em desired system} does not apply to the {\em existing system},
   and suppose it would be violated if applied to the data of the existing system.
   Now, if we try to spin up the desired system on the data set of the existing system, we have a problem.
   Every invariant of the desired system must be satisfied but $u$ isn't.
   So we cannot spin it up.
   In many cases, clearing out all violations before spinning up the desired system might take too much time and new violations will arise in the meantime.
   To solve this problem, we implement every invariant in the migration system initially as a business constraint,
   letting users restore invariance while the new system is running.
   For this purpose, we must define an intermediate system, the \define{migration system},
   which implements the invariants of the desired system as business constraints.
   This migration system must ensure that every violation that is fixed will not return as a violation.
   In this way, the business constraint gradually turns into an invariant.
   Since the number of violations is finite, the business can resolve these violations in finite time.
   New violations that occur are under control of the desired system,
   so the business experiences the migration system as the desired system.
   In this way, the migration system bridges the gap and users get a zero downtime SCDM.

   Summarizing, the following requirements apply to SCDMs
\begin{enumerate}
\item users must experience zero downtime, to enable more frequent SCDMs.
\item users must be able to fix business constraints, so we need an intermediate ``migration system'' to deliver zero downtime.
\item the number of violations that users must fix is finite and decreases (monotonically) over time,
   to ensure that the migration system can be removed after the migration is done.
\end{enumerate}

@@

   Actors (both users and computers) are changing the data in a system continually.
   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
   Events that the system detects may cause the state to change.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.45]{figures/datamigration-Pre-migration.png}
   \end{center}
\caption{Anatomy of an information system}
\label{fig:pre-migration}
\end{figure}
   To keep our theory technology independent, we assume data sets to contain triples.
   This makes our theory valid for any kind of database that triples can represent,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
   An important aspect of the system's semantics are the invariants, sometimes known as integrity rules.
   %The purpose of these rules is to keep all constraints satisfied.

   Ampersand is a compiler that generates information systems from a script.
   It uses a form of heterogeneous relation algebra,
   which works with relations in a way similar to Alloy~\cite{Alloy2006}.
   A system is then described by a set of rules (aka constraint / invariant), which the system keeps satisfied.
   As a result, all invariants are explicit in the source code of a system specified in Ampersand.

@@

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   %Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
   %
   In practice, data migrations are done by deploying the desired system and the existing system side-by-side,
   while transferring data in a controlled fashion, as shown in Figure~\ref{fig:migration phase}.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.35]{figures/datamigration-Migration-phase.png}
   \end{center}
\caption{Migration phase}
\label{fig:migration phase}
\end{figure}
   To automate the migration as much as possible and to achieve zero downtime,
   we also deploy a migration system.
   The migration system has its own schema.
   It comprises the schemas of both the existing system and the desired system,
   so the migration system can copy the data from the existing to the desired system.
   Some of the migration work, however, cannot be automated.
   A migration engineer can minimize manual work by minimizing the size of the increments.
   But data pollution, new business rules, or known issues in the existing system
   may occasionally require tailoring the migration script to specific needs
   that require action of users in production.
   In Ampersand, she can use business constraints to implement these needs.

@@

\section{Terminology}
\label{sct:Terminology}
   An {\em information system} is a combination of data set, schema, and functionality.
   For the purpose of this paper, we ignore functionality captured in user interfaces because it does not impact the migration.
   Section~\ref{sct:Data sets} desribes data sets. Schemas are treated in section~\ref{sct:Schemas}.
   Then section~\ref{sct:Information Systems} defines information systems.

@@

   Relations serve to organize and store data, allowing developers to represent facts.
   In this paper, variables $r$, $s$, and $d$ represent relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
   All relations are taken from an infinite set $\Rels$.
   $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.
   Every relation $r$ has a name, a source concept, and a target concept.
   The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.
   The part $\pair{A}{B}$ is called the {\em signature} of the relation.

@@

   A relation $r$ can serve as a container of pairs,
   as defined by the function $\popF{r}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
   It defines a set of pairs, also known as the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
\label{eqn:pop-rel}
\end{equation}
%   Equation~\ref{eqn:wellTypedEdge} implies that for every data set $\dataset$:
%\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\instance_{\dataset}A\ \wedge\ b\instance_{\dataset}B\]
%   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
%
   We overload the notation $\popF{}$ so we can use it on concepts $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
   and expressions. We also define the difference of populations, both for relations and concepts:
\begin{align}
   \pop{C}{\dataset}&= \{ x\mid\ x\ \instance_{\dataset}\ C\}
\label{eqn:pop-concept}\\
   \pop{x-y}{\dataset}&= \pop{x}{\dataset} - \pop{y}{\dataset}
\label{eqn:pop-expr}
\end{align}

\subsection{Violations}
\label{sct:Violations}
   To satisfy a constraint $u$ means that it is true for all triples in the data set.
   Let us assume that we can represent each violation of a constraint $u$ as a pair
   and let the function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produce these violations.
   Note that this assumption implies that rules have a signature, just like a relations.
   So, we can represent an invariant $u$ as a triple $\triple{\violC{u}}{A}{B}$
   that satisfies:
\begin{equation}
   \pair{a}{b}\in\viol{\triple{\violC{u}}{A}{B}}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
\end{equation}
   We represent an enforce rule $u$ as a pair $\pair{\declare{n}{A}{B}}{\violC{u}}$,
   where relation $\declare{n}{A}{B}$ is used to restore invariance of the enforce rule.
   The enforce rule satisfies:
\begin{equation}
   \pair{a}{b}\in\viol{u}{\dataset}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B\ \wedge\ \viol{u}{\dataset\cup\triple{a}{\declare{n}{A}{B}}{b}}=\emptyset
\end{equation}

\subsection{Schemas}
\label{sct:Schemas}
   Schemas serve to capture the semantics of a system~\cite{Spivak2012}.
   It defines concepts, relations, and invariants.
   We assume that a software engineer defines a schema on design time, and a compiler checks whether the semantics are consistent.
   Error messages from the compiler are prohibitive for generating code,
   which prevents a substantial class of mistakes to ever reach end users.

   We describe a schema $\schema$ via a tuple $\quadruple{\concepts}{\rels}{\rules}{\enforces}$,
   in which $\concepts\subseteq \Concepts$ is a finite set of concepts,
   $\rels\subseteq \Rels$ is a finite set of relations,
   $\rules\subseteq \Triple{(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}})}{\Concepts}{\Concepts}$ is a finite set of invariants,
   and $\enforces\subseteq \Pair{\Rels}{\left(\Dataset \to \powerset{\Pair{\Atoms}{\Atoms}}\right)}$ is a finite set of enforcement rules.

   Each invariant in a schema serves to constrain the data set at runtime, to ensure its semantic integrity.
   Every invariant is an element of an infinite set called $\Rules$.
   In this paper, variables $u$ and $v$ represent rules.
   A invariant $u$ in the schema is a triple $\triple{\violC{u}}{A}{B}$ such that $\viol{u}{\dataset}$ represents any violations to $u$ in data set $\dataset$.
   In Ampersand, rules are well-typed, just like relations:
   Every violation $\pair{a}{b}$ of a invariant $\triple{\violC{u}}{A}{B}$ must satisfy $a\instance A$ and $b\instance B$.
   % For every data set $\dataset$, the set $\viol{u}{\dataset}$ is a subset of $\pop{A}{\dataset}\times\pop{B}{\dataset}$.
   If there are no violations, that is: $\viol{u}{\dataset} = \emptyset{}$, then we say that $u$ is satisfied.

@@

   A rule, {\tt EnrollRule} completes the script.
   It states that a student can enroll for any module that is part of a course she takes.
   % In Ampersand, which is a syntactically sugared form of relation algebra~\cite{JoostenRAMiCS2017},
   % each rule has a name and each rule has a role to maintain its invariance:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
\end{verbatim}
   The Ampersand compiler defines $\viol{\tt EnrollRule}{\dataset}$ as
\begin{equation}
   \pop{\tt isEnrolledFor}{\dataset}-\{\pair{s}{m} \mid\exists c.~s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c\}
\end{equation}
   This rule is satisfied if $\viol{\tt EnrollRule}{\dataset}=\emptyset$, which means that
\begin{equation}
   \begin{array}{l}
   \forall \pair{s}{m}\in\pop{\tt isEnrolledFor}{\dataset}.~ \exists c\in\text{\tt Course}~.\\
   s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
   \end{array}
\label{eqn:example isEnrolledFor}
\end{equation}
   In this example, relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty, so the rule is satisfied.

@@

   The MoC is reached for data set $\dataset'$ whenever $\pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'}=\emptyset$ for all rules.
   This should happen after a finite number of inserting events: let $I = \{{\tt fixed}_u \mid u\in\rules_{\schema'}\}$ be the relations that keep track of which violations have been fixed, then there is no infinite chain of systems $\migrsys' = \migrsys_0, \migrsys_1,\ldots$ such that $\migrsys_i\xrightarrowdbl[I]{\overrightarrow\rels_{\schema'}}\migrsys_{i+1}$.
   This shows that when users get rid of violations, there is real progress: they can only do this a finite number of times.
   Next we ask ourselves: does this mean we always reach the MoC?

@@

   For this last question, we need to add a condition:
   There needs to be a system that has the desired schema $\schema'$.
   If rules in $\schema'$ are such that they cannot all be satisfied, regardless of the data set, no such system exists and we have no hope of reaching it.
   We assume that the rules of $\schema'$ are intended to prevent data pollution, such that a data set $\dataset'$ describing the `real world' would be one that satisfies them.
   So let $\pair{\dataset'}{\schema'}$ be a system, then we can get to it through our migration system.
   To show this, we construct a migration system $\migrsys^C$ satisfying $\pop{{\tt fixed}_u}{\dataset_{\migrsys^C}}=\pop{{\tt viols}_u}{\dataset_{\migrsys^C}}$ and $\migrsys'\xrightarrow[I]{\overrightarrow\rels_{\schema'}} \migrsys^C$.
   We can do so by simply stating its triples, since the schema is determined by $\migrsys'$:

@@

   This work does not consider what to do about (user) interfaces.
   Instead, it models events by assuming that any change to the data set can be achieved.
   In practice, such changes need to be achieved through interfaces.
   Most Ampersand systems indeed allow the users of the system to edit the data set quite freely through the interfaces.
   However, some interfaces may require certain constraints to be satisfied, which means that interfaces of the desired system might break.
   In the spirit of the approach outlined here, we hope to generate migration interfaces that can replace any broken interfaces until the Moment of Transition.
   How to do this is future work.
